Supporting Information for\\ ``A weakly structured stem for human origins in Africa''

### 2022-01-09 paper.tex
edits to figures and results

We inferred detailed demographic histories using 4x-8x whole-genome sequencing
data for four diverse African populations, comprising the Nama (Khoe-San from
South Africa, newly presented here), Mende (MSL from Sierre Leone, Phase 3 of
the 1000 Genomes Project \citep{1000_Genomes_Project_Consortium2015-zq}), Gumuz
(recent descendents of a hunter-gatherer group from Ethiopia
\citep{Gurdasani2015-qy,Gopalan2019-wd}), and Eastern African agriculturalists
(Amhara and Oromo from Ethiopia \citep{Gurdasani2015-qy}). The Amhara and Oromo
populations, despite speaking distinct Afro-Asiatic languages, are highly
genetically similar \citep{Pagani2015-pz,Gopalan2019-wd} and thus the two
groups were combined for a larger sample size (Figure~\ref{fig:1}). We also
included the British (GBR) from the 1000 Genomes Project in our demographic
models as a representative source of back-to-Africa gene flow and recent
colonial admixture in South Africa. Finally, we added a high-coverage ancient
Neanderthal genome from Vindija Cave, Croatia \citep{Prufer2017-kk} to account
for archaic gene flow from Neanderthals in non-Africans and gauge the relative
time depth of divergence, assuming Neanderthals diverged 550kya from a common
stem. For each population, we computed low-order allele frequency and linkage
disequilibrium (LD) statistics that are well suited for both low- and
high-coverage genomes \citep{Ragsdale2019-nt,Ragsdale2020-nz}. Using a
maximum-likelihood inference framework, we then fit to these statistics a
family of parameterized demographic models that involve population splits, size
changes, continuous and variable migration rates, and punctuated admixture
events, to learn about the nature of population structure over the past million
years.

@@

We began with a model of geographic expansion from a single ancestral source
followed by migration between populations
(Figure~\ref{fig:supp-possible-models}A), without allowing for contribution
from an archaic African lineage. As expected \citep{Ragsdale2019-nt}, this
first model was a poor fit to the data qualitatively
(Figure~\ref{fig:supp-single-origin-fits}) and quantitatively (log-likelihood
($LL$) $\approx -189,000$, Table~\ref{tab:supp-single-origin}). We next explored a
suite of models in which population structure predates the differentiation of
contemporary groups, including models allowing for ancestral reticulation
(Figure~\ref{fig:supp-possible-models}B), archaic admixture
(Figure~\ref{fig:supp-possible-models}C), and African multi-regionalism
(Figure~\ref{fig:supp-possible-models}D).

Regardless of the model choice for early epochs, inference of human
demographic history in the last 150kya was remarkably robust. The earliest divergence
among contemporary sampled human populations differentiates the southern
African Nama from other African groups between 110--135kya, with low to moderate
levels of subsequent gene flow \aprcomment{(Table 1)}. In none of the
high-likelihood models which we explored did the divergence between Nama and
other populations exceed $\sim$150kya. We conclude that, regardless of model
specification, geographic patterns of \emph{Homo sapiens} population structure
date back to the Late Middle Stone Age in Africa, likely arising during MIS 5.
Although we find evidence for earlier population structure in Africa (see
below), present-day populations cannot be easily mapped onto these more ancient
stem groups as only a small proportion of drift between modern populations can
be attributed to drift between stems \aprcomment{(Figure 3)}.

Given this consistency in inferred recent history and the numerical challenge
of optimizing a large number of parameters, we fixed several parameters related
to recent population history so as to focus on more ancient events (Supp.
Methods). The time of divergence of Western and Eastern African populations was
set to 60kya, just prior to the split of Eurasians and East Africans 50kya. Both
models predict relatively high subsequent gene flow between the two regions
($m=2\times10^{-4}$). Admixture from Neanderthals to Europeans directly
following the out-of-Africa migration was set to 1.5\% at 45kya (Supp.
Methods). Back-to-Africa gene flow at the beginning of the Holocene primarily
affected the ancestors of the Ethiopian agricultural populations, comprising
over half of their genetic ancestry, estimated to be 64--65\%. The past 5,000
years also saw major demographic changes, including strong population growth
for Western Africans as they specialized in yam and oil palm agriculture
(estimated 3-fold growth). We observe significant gene flow from the Amhara and
Oromo into the Nama, a signal which is likely a proxy for the movement of
Eastern African caprid and cattle pastoralists
\citep{Henn2008-xo,Breton2014-xb}, here estimated to constitute a 25\% ancestry
contribution 2,000 ya. Colonial period admixture from Europeans into the Nama
was estimated at 15\%, similar to proportions inferred by ADMIXTURE
\aprcomment{cite} (Figure~\ref{fig:1}).

@@

To account for population structure prior to 135kya, we allowed for two or more
“stem” populations which could diverge either before or after the Neanderthal
split. We considered models both with and without migration between these stem
populations, and in both cases we tested two different types of gene exchange
during the expansion phase. 1) One of the stem population expands (splits into
contemporary populations), and the other stem population(s) has continuous
symmetric migration with those populations; or 2) expansion of one or more of
the stem populations, with an instantaneous pulse (or “merger”) from the other
stem population. These models are meant to capture scenarios 2B and 2C (Figure
2). \aprcomment{update this last sentence}

\begin{table}[t!]
    \centering
    \scriptsize
    \begin{tabular}{lllrrrr}
        \toprule
        \textbf{Likelihood} & \textbf{Label} & \textbf{Population Pair} &
            \textbf{Divergence} & \textbf{Migration rate} & \textbf{Migration} \\
        & & & 
            \textbf{Time (kya)} & \textbf{per generation} & \textbf{duration (ky)} \\
        \midrule
        \textbf{Continuous Model} & & & & & \\
        LL= -115,500 & a & Stem 1, Stem 2 & 1,163 & 6.43e-5 & 1,028 \\
        & b & Stem 2, Nama & NA & 5.82e-5 & 130 \\
        & c, d & Stem 2, Other Africans & NA & 3.10e-5, \textbf{1.64e-4} & 130, 55 \\
        & e, f & Nama, Other Africans* & 135 & 4.1e-5, 9.8e-6 & 135, 60 \\
        & g & Mende, East Africans & 60 & \textbf{2.14e-4} & 60 \\
        & h & East Africans, British & 50 & 4.17e-5 & 50 \\
        & i & Gumuz, Amhara/Oromo & 12 &\textbf{3.36e-4} & 12 \\
        \textbf{Merger Model} & & & & & \\
        LL= -102,600 & a & Stem 1, Stem 2 & 1,442 & \textbf{1.16e-4} & 963 \\
        & -- & Stem 1S, Stem 1E & 479 & 0 (fixed) & -- \\
        & b & Stem 2 to Nama & 119 & \textbf{0.71} & pulse \\
        & c & Stem 2 to Stem 1E & 98 & \textbf{0.50} & pulse \\
        & d & Stem 2 to Mende & 25 & \textbf{0.18} & pulse \\
        & e, f & Nama, Other Africans & 119 & 4.4e-5, 7.1e-6 & 119, 60 \\
        & g, & Mende, East Africans & 60 & \textbf{1.98e-4} & 60 \\
        & h & East Africans, British & 50 & 3.87e-5 & 50 \\
        & i & Gumuz, Amhara/Oromo &12 & \textbf{3.59e-4} & 12 \\
        \bottomrule
    \end{tabular}
    \caption{
        \textbf{Migration and divergence parameters from best fit models.}\\
        \aprcomment{Labels are keyed to Figure~\ref{fig:2}}
    }
    \label{tab:migration-rates}
\end{table}

Allowing for continuous migration between the stem populations substantially
improved the fits relative to zero migration (e.g. $LL \approx -102,600$ vs.
$-107,700$). With continuous migration, population structure extends back to
1.1--1.4Mya \aprcomment{(Table 1)}. Migration between the stems in these models
is low but non-negligible, with a fraction of migrant lineages each generation
estimated as $m=6.43\times10^{-5}-1.16\times10^{-4}$. For comparison, this is
at least as large as migration inferred between the Ethiopian East Africans and
British populations over the past 50kya \aprcomment{(Table 1)}. This ongoing
(or, at least, periodic) gene flow qualitatively distinguishes these models
from previously proposed archaic admixture models \aprcomment{(Figure S1C)} as
the early branches remain closely related throughout (Figure 3).
\aprcomment{this last phrase may need to be updated, with regard to Fig
\ref{fig:supp-FST-panels}}

Likelihood-based parameter confidence intervals  \sgcomment{or: Bootstrap-based confidence interval reflecting the finite length of the genome?}   are relatively narrow within
each specified model
(Tables~\ref{tab:supp-single-origin}--\ref{tab:supp-merger-with-stem-migration}),
reflecting an informative statistical approach. As expected, model assumptions
have a larger impact on parameter estimates. To convey model uncertainty, we
highlight features of two models with the highest \aprcomment{not the
highest\ldots merger-without-stem-migration at -107,650}log likelihoods. These
differ primarily in whether early stems are assumed to give rise to modern
populations in multiple discrete merger events or in continuous migrations
\aprcomment{(Figure 2)}. Additional models are discussed in
\aprcomment{Supplement XXX}.

@@

Under the continuous-migration model, the two stems give rise to contemporary
populations in western, southern and eastern Africa but with differing
proportions of ancestry derived from “Stem 1” and “Stem 2” (Figure 2). For
example, the Nama and Gumuz have lower levels of migration with Stem 2
($m=5.8\times10^{-5}$ and $3.1\times10^{-5}$, respectively), relative to the
Mende. The Mende continue to exchange migrants with Stem 2
($m=1.6\times10^{-4}$) long after their separation 60kya; this is a similar
level of gene flow as between the Mende and Eastern Africans (Table 1). 
\aprcomment{paragraphs above need work, and condensing}

\aprcomment{also below, too much information - we don't actually believe these
models}

@@

\begin{figure}[t!]
    \centering
    \includegraphics{figures/fig3.pdf}
    \caption{
        \textbf{fig3}
    }
    \label{fig:3}
\end{figure}

@@

\begin{figure}[ht]
\begin{center}
    \makebox[\textwidth][c]{} % \includegraphics{figures/fig3.pdf}}
\caption{\textbf{The fourth main figure}.
    A placeholder - predictions (FST and/or f4).
}
\label{fig:4}
\end{center}
\end{figure}

>>>>

We inferred detailed demographic histories using 4x-8x whole-genome sequencing
data for four diverse African populations, comprising the Nama (Khoe-San from
South Africa, newly presented here), Mende (from Sierre Leone, MSL from the
Phase 3 1000 Genomes Project \citep{1000_Genomes_Project_Consortium2015-zq}),
Gumuz (recent descendents of a hunter-gatherer group from Ethiopia
\citep{Gurdasani2015-qy,Gopalan2019-wd}), and Eastern African agriculturalists
(Amhara and Oromo from Ethiopia \citep{Gurdasani2015-qy}). The Amhara and Oromo
populations, despite speaking distinct Afro-Asiatic languages, are highly
genetically similar \citep{Pagani2015-pz,Gopalan2019-wd} and thus the two
groups were combined for a larger sample size (Figure~\ref{fig:1}). We also
included the British (GBR) from the 1000 Genomes Project in our demographic
models as a representative source of back-to-Africa gene flow and recent
colonial admixture in South Africa. Finally, we used a high-coverage ancient
Neanderthal genome from Vindija Cave, Croatia \citep{Prufer2017-kk} to account
for archaic gene flow from Neanderthals in non-Africans and gauge the relative
time depth of divergence, assuming Neanderthals diverged 550kya from a common
stem. For each population, we computed low-order allele frequency and linkage
disequilibrium (LD) statistics that are well suited for both low- and
high-coverage genomes \citep{Ragsdale2019-nt,Ragsdale2020-nz}. Using a
maximum-likelihood inference framework, we then fit to these statistics a
family of parameterized demographic models that involve population splits, size
changes, continuous and variable migration rates, and punctuated admixture
events, to learn about the nature of population structure over the past million
years.

@@

We began with a model of geographic expansion from a single ancestral source
followed by migration between populations
(Figure~\ref{fig:supp-possible-models}A), without allowing for contribution
from an archaic African lineage or population structure prior to the expansion.
As expected \citep{Ragsdale2019-nt}, this first model was a poor fit to the
data qualitatively (Figure~\ref{fig:supp-single-origin-fits}) and
quantitatively (log-likelihood ($LL$) $\approx -189,400$,
Table~\ref{tab:supp-single-origin}). We next explored a suite of models in
which population structure predates the differentiation of contemporary groups,
including models allowing for ancestral reticulation
(Figure~\ref{fig:supp-possible-models}B), archaic admixture
(Figure~\ref{fig:supp-possible-models}C), and African multi-regionalism
(Figure~\ref{fig:supp-possible-models}D).

Regardless of the model choice for early epochs, inference of human
demographic history in the last 150kya was remarkably robust. The earliest divergence
among contemporary sampled human populations differentiates the southern
African Nama from other African groups between 110--135kya, with low to moderate
levels of subsequent gene flow (Table\ref{tab:migration-rates}). In none of the
high-likelihood models which we explored did the divergence between Nama and
other populations exceed $\sim$140kya. We conclude that, regardless of model
specification, geographic patterns of \emph{Homo sapiens} population structure
date back to the Late Middle Stone Age in Africa, likely arising during MIS 5.
Although we find evidence for earlier population structure in Africa (see
below), present-day populations cannot be easily mapped onto these more ancient
stem groups as only a small proportion of drift between modern populations can
be attributed to drift between stems (Figures~\ref{fig:3} and
\ref{fig:supp-f4s-single-origin}--\ref{fig:supp-f4s-merger-with-stem-migration}).

Given this consistency in inferred recent history and the numerical challenge
of optimizing a large number of parameters, we fixed several parameters related
to recent population history so as to focus on more ancient events (Supp.
Methods). The time of divergence of Western and Eastern African populations was
set to 60kya, just prior to the split of Eurasians and East Africans 50kya. All
models predict relatively high subsequent gene flow between the two regions
($m\approx2\times10^{-4}$). Admixture from Neanderthals to Europeans directly
following the out-of-Africa migration was set to 1.5\% at 45kya (Supp.
Methods). Back-to-Africa gene flow at the beginning of the Holocene primarily
affected the ancestors of the Ethiopian agricultural populations, comprising
over half of their genetic ancestry, estimated to be 64--65\%. The past 5,000
years also saw major demographic changes, including strong population growth
for Western Africans as they specialized in yam and oil palm agriculture
(estimated 3-fold growth). We observe significant gene flow from the Amhara and
Oromo into the Nama, a signal which is likely a proxy for the movement of
Eastern African caprid and cattle pastoralists
\citep{Henn2008-xo,Breton2014-xb}, here estimated to constitute a 25\% ancestry
contribution 2,000 ya. Colonial period admixture from Europeans into the Nama
was estimated at 15\%, similar to proportions inferred by \texttt{ADMIXTURE}
\citep{Alexander2009-sw} (Figure~\ref{fig:1}).

@@

To account for population structure prior to 135kya, we allowed for two or more
“stem” populations which could diverge either before or after the Neanderthal
split. We considered models both with and without migration between these stem
populations, and in both cases we tested two different types of gene exchange
during the expansion phase. 1) One of the stem population expands (splits into
contemporary populations), and the other stem population(s) has continuous
symmetric migration with those populations; or 2) one or more of the stem
populations expands, with instantaneous pulse (or “merger”) events from the
other stem population, so that diversifying branches are formed through the
coalescence of multiple ancestral populations. Depending on parameter values,
this scenario encompasses archaic introgression and
fragmentation-and-coalescence models.

\begin{table}[t!]
    \centering
    \scriptsize
    \begin{tabular}{lllrrrr}
        \toprule
        \textbf{Likelihood} & \textbf{Label} & \textbf{Population Pair} &
            \textbf{Divergence} & \textbf{Migration rate} & \textbf{Migration} \\
        & & & 
            \textbf{Time (kya)} & \textbf{per generation} & \textbf{duration (ky)} \\
        \midrule
        \textbf{Continuous Model} & & & & & \\
        LL= -115,500 & a & Stem 1, Stem 2 & 1,163 & 6.43e-5 & 1,028 \\
        & b & Stem 2, Nama & NA & 5.82e-5 & 130 \\
        & c, d & Stem 2, Other Africans* & NA & 3.10e-5, \textbf{1.64e-4} & 130, 55 \\
        & e, f & Nama, Other Africans* & 135 & 4.1e-5, 9.8e-6 & 135, 60 \\
        & g & Mende, East Africans & 60 & \textbf{2.14e-4} & 60 \\
        & h & East Africans, British & 50 & 4.17e-5 & 50 \\
        & i & Gumuz, Amhara/Oromo & 12 &\textbf{3.36e-4} & 12 \\
        \textbf{Merger Model} & & & & & \\
        LL= -102,600 & a & Stem 1, Stem 2 & 1,442 & \textbf{1.16e-4} & 963 \\
        & -- & Stem 1S, Stem 1E & 479 & 0 (fixed) & -- \\
        & b & Stem 2 to Nama & 119 & \textbf{0.71} & pulse \\
        & c & Stem 2 to Stem 1E & 98 & \textbf{0.50} & pulse \\
        & d & Stem 2 to Mende & 25 & \textbf{0.18} & pulse \\
        & e, f & Nama, Other Africans* & 119 & 4.4e-5, 7.1e-6 & 119, 60 \\
        & g, & Mende, East Africans* & 60 & \textbf{1.98e-4} & 60 \\
        & h & East Africans, British & 50 & 3.87e-5 & 50 \\
        & i & Gumuz, Amhara/Oromo &12 & \textbf{3.59e-4} & 12 \\
        \bottomrule
    \end{tabular}
    \caption{
        \textbf{Migration and divergence parameters from best fit models.}
        Labeled migration rates correspond to symmetric continuous migration
        bands shown in Figure~\ref{fig:2}. Both the continuous-migration and
        merger models \ldots
        \aprcomment{also define *}
    }
    \label{tab:migration-rates}
\end{table}

Allowing for continuous migration between the stem populations substantially
improved the fits relative to zero migration (e.g. $LL \approx -102,600$ vs.
$-107,700$ in the merger scenario and $LL \approx -115,500$ vs. $-128,000$ with
continuous migration). With continuous migration between stems, population
structure extends back to 1.1--1.4Mya (Table~\ref{tab:migration-rates}).
Migration between the stems in these models is moderate, with a fraction of
migrant lineages each generation estimated as
$m=6.43\times10^{-5}-1.16\times10^{-4}$. For comparison, this roughly compares
to inferred migration rates between connected contemporary populations over the
past 50kya (Table~\ref{tab:migration-rates}). This ongoing (or, at least,
periodic) gene flow qualitatively distinguishes these models from previously
proposed archaic admixture models (Figure~\ref{fig:supp-proposed-models}C) as
the early branches remain closely related (Figure 3). \aprcomment{this last
phrase may need to be updated, with regard to Fig \ref{fig:supp-FST-panels}}

Bootstrap-based confidence intervals are relatively narrow within each
specified model
(Tables~\ref{tab:supp-single-origin}--\ref{tab:supp-merger-with-stem-migration}),
reflecting an informative statistical approach. As expected, model assumptions
have a larger impact on parameter estimates. To convey model uncertainty, we
highlight features of two models with high likelihoods. These differ primarily
in whether early stems are assumed to give rise to modern populations in
multiple discrete merger events or in continuous migrations
(Figure~\ref{fig:2}). Additional models are discussed in the Supporting
Material.

@@

Under the continuous-migration model, the two stems give rise to contemporary
populations in western, southern and eastern Africa but with differing
proportions of ancestry derived from “Stem 1” and “Stem 2” (Figure 2). For
example, the Nama and Gumuz have lower levels of migration with Stem 2
($m=5.8\times10^{-5}$ and $3.1\times10^{-5}$, respectively), relative to the
Mende. The Mende continue to exchange migrants with Stem 2
($m=1.6\times10^{-4}$) long after their separation 60kya; this is a similar
level of gene flow as between the Mende and Eastern Africans (Table 1).

@@

\begin{figure}[t!]
    \centering
    %\includegraphics{figures/fig3.pdf}
    \caption{
        \textbf{fig3}
    }
    \label{fig:3}
\end{figure}

@@

\begin{figure}[ht!]
    \centering
    \includegraphics{figures/fig4}
    \caption{fig4}
    \label{fig:4}
\end{figure}

### 2022-01-09 supplement.tex
edits to figures and results

\begin{figure}[ht]
    \centering
    \includegraphics[width=\textwidth]{figures/supp-fsts-all-models.pdf}
    \caption{
        \textbf{Predicted \bm{$F_{ST}$} between contemporaneous populations over time.}
        \aprcomment{Plot ordering, line colors and styles, etc need to be cleaned up.
        Add panel labels.}
        All models (A: single origin, B: continuous migration, C: merger without stem
        migration, D: merger with stem migration) match observed $F_{ST}$ between
        sampled populations. The continuous-migration model predicts that human stem
        populations remain genetically similar, while the merger with and without
        stem migration models predict a period of increased $F_{ST}$ between ancestors
        of modern humans. This is largely due to the very small inferred $N_e$ in one of
        the stem branches, which leads to a rapid increase in differentiation due to
        drift.
    }
    \label{fig:supp-FST-panels}
\end{figure}

\begin{figure}[ht]
    \centering
    \includegraphics[width=\textwidth]{figures/supp-h12s-all-models.pdf}
    \caption{
        \textbf{Predicted pairwise differentiation between contemporaneous populations.}
        \aprcomment{Plot ordering, line colors and styles, etc need to be cleaned up.
        Add panel labels.}
        $H_{i,j}$ is the predicted pairwise differentiaion between two populations,
        which we normalize by the observed pairwise diversity in the Mende sample.
        The single-origin model (A) shows larger deviations between observed and
        expected differentiation than the models with early human structure
        (B: continuousmigration, C: merger without stem migration, D: merger with 
        stem migration). In the merger models (C and D), while $F_{ST}$ is large
        between early stems due to the sharp bottleneck in one of those branches,
        the pairwise differences are comparable to the continuous-migration model.
        Stem 1E and Stem 1S have low differentiation, as the bottleneck in Stem 1
        (which they both split from) sharply reduces diversity by the time of the
        split. This has the effect of large $F_{ST}$ but small $H_{i,j}$.
    }
    \label{fig:supp-h12-panels}
\end{figure}

>>>>

\begin{figure}[ht]
    \centering
    \includegraphics[width=\textwidth]{figures/supp-fsts-all-models.pdf}
    \caption{
        \textbf{Predicted \bm{$F_{ST}$} between contemporaneous populations over time.}
        All models (A: single origin, B: continuous migration, C: merger without stem
        migration, D: merger with stem migration) match observed $F_{ST}$ between
        sampled populations. The continuous-migration model predicts that human stem
        populations remain genetically similar, while the merger with and without
        stem migration models predict a period of increased $F_{ST}$ between ancestors
        of modern humans. This is largely due to the very small inferred $N_e$ in one of
        the stem branches, which leads to a rapid increase in differentiation due to
        drift.
    }
    \label{fig:supp-FST-panels}
\end{figure}

\begin{figure}[ht]
    \centering
    \includegraphics[width=\textwidth]{figures/supp-h12s-all-models.pdf}
    \caption{
        \textbf{Predicted pairwise differentiation between contemporaneous populations.}
        $H_{i,j}$ is the predicted pairwise differentiaion between two populations,
        which we normalize by the observed pairwise diversity in the Mende sample.
        The single-origin model (A) shows larger deviations between observed and
        expected differentiation than the models with early human structure
        (B: continuousmigration, C: merger without stem migration, D: merger with 
        stem migration). In the merger models (C and D), while $F_{ST}$ is large
        between early stems due to the sharp bottleneck in one of those branches,
        the pairwise differences are comparable to the continuous-migration model.
        Stem 1E and Stem 1S have low differentiation, as the bottleneck in Stem 1
        (which they both split from) sharply reduces diversity by the time of the
        split. This has the effect of large $F_{ST}$ but small $H_{i,j}$.
    }
    \label{fig:supp-h12-panels}
\end{figure}
